% !TEX root = femlecture.tex
\chapter{算子理论}

% Sobolev spaces are fundamental in the study of partial differential equations and their
% numerical approximations. In this chapter, we shall give brief discussions on the Sobolev
% spaces and the regularity theory for elliptic boundary value problems.


\section{Preliminary From Functional Analysis}


Let $X$ and $Y$ be two Banach spaces in this section. See \cite{Brezis2011, Weidmann1980} for details in this section.

\subsection{Closed operators}
\begin{defn}[Unbounded linear operator]
Let $X$ and $Y$ be two Banach spaces. An {\bf unbounded linear operator} from
$X$ into $Y$ is linear map $A: D(A)\subset X\to Y$ defined on a linear subspace $D(A)\subset X$ with values in $Y$. The set $D(A)$ is called the domain of $A$.
\end{defn}

\begin{defn}[Graph]
Let $X$ and $Y$ be two Banach spaces. Let $A: D(A)\subset X\to Y$ be a linear operator.
Define the \textbf{graph} of $A$ as the following linear subspace
\[
\{(x, Ax)\in X\times Y: X\in D(A)\}.
\]
The linear operator $A$ is said to be \textbf{closed} if the graph of $A$ is closed in $X\times Y$.
\end{defn}

\begin{lemma}
The linear operator $A: D(A)\subset X\to Y$ is closed if and only if \textcolor[rgb]{1.00,0.00,0.00}{its domain $D(A)$ is a complete space with respect to the graph norm $\sqrt{\|x\|_X^2+\|Ax\|_{Y}^2}$}.
\end{lemma}

\begin{lemma}
If the linear operator $A: D(A)\subset X\to Y$ is closed, then \textcolor[rgb]{1.00,0.00,0.00}{$N(A)$ is closed; however, $R(A)$ need not be closed}.
\end{lemma}

\begin{theorem}[Closed graph theorem]
Let $X$ and $Y$ be two Banach spaces. If $A: D(A)\subset X\to Y$ is a closed linear operator and \textcolor[rgb]{1.00,0.00,0.00}{$D(A)$ is closed}, then \textcolor[rgb]{1.00,0.00,0.00}{$A$ is continuous (bounded)}.
\end{theorem}



\subsection{Adjoints in Banach spaces}
\begin{defn}[Adjoint $A'$]
Suppose $A: D(A)\subset X\to Y$ be a linear operator that is densely defined. Let
\[
D(A')=\{y'\in Y': \exists x'\in X' \textrm{ such that } \langle y', Ax\rangle=\langle x', x\rangle\quad \forall x\in D(A)\},
\]
where $X'$ and $Y'$ are the dual spaces of $X$ and $Y$ respectively.
Then we define $A'y':=x'$. The linear operator
\[
A': D(A')\subset Y'\to X'
\]
is called the \textbf{adjoint} of $A$. The \textbf{domain} $D(A')$ can be rewritten as
\[
D(A')=\{y'\in Y': \exists c\geq0 \textrm{ such that } |\langle y', Ax\rangle|\leq c\|x\|_X\quad \forall x\in D(A)\}.
\]
The fundamental relation between $A$ and $A'$ is given by
\[
\langle y', Ax \rangle_{Y'\times Y}=\langle A'y', x \rangle_{X'\times X}\quad\quad\forall x\in D(A),\quad y'\in D(A').
\]
\end{defn}

\begin{lemma}
If $A: X\to Y$ is a linear bounded operator then $A': Y'\to X'$ is also a linear bounded operator and, moreover,
\[
\|A'\|_{\mathcal L(Y', X')}=\|A\|_{\mathcal L(X, Y)}.
\]
\end{lemma}


\begin{lemma}
Let $A: D(A)\subset X\to Y$ be a densely defined linear operator. Then \textcolor[rgb]{1.00,0.00,0.00}{the adjoint $A': D(A')\subset Y'\to X'$ is closed}.
\end{lemma}

\begin{theorem}
Let $X$ and $Y$ be two \textcolor[rgb]{1.00,0.00,0.00}{reflexive} Banach spaces. Let $A: D(A)\subset X\to Y$ be an unbounded linear operator that is densely defined and closed. Then \textcolor[rgb]{1.00,0.00,0.00}{$D(A')$ is dense in $Y'$}. Thus $A''$ is well defined ($A'': D(A'')\subset X''\to Y''$) and it may
also be viewed as an unbounded operator from $X$ into $Y$. Then we have
\[
A''=A.
\]
\end{theorem}

\begin{defn}
If $M\subset X$ is a linear subspace,  we define the annihilator of $M$ in $X'$
\[
M^{\perp}=\{f\in X': \langle f, x\rangle=0\quad \forall x\in M\}.
\]
If $N\subset X'$ is a linear subspace we set
\[
N^{\perp}=\{x\in X: \langle f, x\rangle=0\quad \forall f\in N\}.
\]
\end{defn}

\begin{lemma}
Let $M\subset X$ be a linear subspace. Then
\[
(M^{\perp})^{\perp}=\overline{M}.
\]
Let $N\subset X'$ be a linear subspace. Then
\[
(N^{\perp})^{\perp}\supset\overline{N}.
\]
Moreover, if $X$ is reflexive, then
\[
(N^{\perp})^{\perp}=\overline{N}.
\]
\end{lemma}

\begin{theorem}
Let $A: D(A)\subset X\to Y$ be an unbounded linear operator that is densely defined and closed. Then
\[
N(A)=R(A')^{\perp},
\quad
N(A')=R(A)^{\perp},
\]
\[
N(A)^{\perp}\supset \overline{R(A')},
\quad
N(A')^{\perp}= \overline{R(A)}.
\]
Moreover, if $X$ is reflexive, then
\[
N(A)^{\perp}= \overline{R(A')}.
\]
\end{theorem}

\begin{theorem}[Closed range theorem]
Let $A: D(A)\subset X\to Y$ be an unbounded linear operator that is densely defined and closed. The following properties are equivalent:
\begin{enumerate}[(i)]
\item $R(A)$ is closed.
\item $R(A')$ is closed.
\item $R(A)=N(A')^{\perp}$.
\item $R(A')=N(A)^{\perp}$.
\end{enumerate}
\end{theorem}

\begin{theorem}
Let $A: D(A)\subset X\to Y$ be a closed unbounded linear operator. Then $R(A)$ is closed if and only if there exists a constant $C$ such that
\[
\textrm{dist}(x, N(A))\leq C\|Ax\|_Y\quad \forall x\in D(A).
\]
\end{theorem}

\begin{theorem}
Let $A: D(A)\subset X\to Y$ be an unbounded linear operator that is densely defined and closed. The following properties are equivalent:
\begin{enumerate}[(a)]
\item $A$ is surjective, i.e. $R(A)=Y$,
\item there is a constant $C$ such that
\[
\|y\|_{Y'}\leq C\|A'y\|_{X'}\quad \forall y\in D(A'),
\]
\item $N(A')=\{0\}$ and $R(A')$ is closed.
\end{enumerate}
\end{theorem}

\begin{theorem}
Let $A: D(A)\subset X\to Y$ be an unbounded linear operator that is densely defined and closed. The following properties are equivalent:
\begin{enumerate}[(a)]
\item $A'$ is surjective, i.e. $R(A')=X'$,
\item there is a constant $C$ such that
\[
\|x\|_{X}\leq C\|Ax\|_{Y}\quad \forall x\in D(A),
\]
\item $N(A)=\{0\}$ and $R(A)$ is closed.
\end{enumerate}
\end{theorem}








